Mining of Keystroke and Mouse Dynamics to Increase the Engagement of Students with Programming Assignments

### 2018-04-19 paper.tex
Improves bibliography
Extends introduction

\section{Introduction}
Programming courses are often regarded as difficult
\cite{robins2003learning, lahtinen2005study}
and there is a common conception that they often have a high
failure rate \cite{bennedsen2007failure}.
Jenkins \cite{jenkins2001motivation} argues that this lack of success has
multiple factors:
\begin{itemize}
\item There is a deep relation with the expectations, attitudes,
and previous experiences of the teaching staff and students.
\item The nature of the subject that involves the
learning of new abstract constructs, syntax and tools.
\item Groups are often heterogeneous and thus it is difficult to design courses
that are beneficial for everyone. Many students begin to program when they
are at their first year of university which means they are tackling a totally
new topic that does not respond to their habitual study approaches.
\end{itemize}

Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is
very problem-solving intensive, which means that a high precision is required
because even the slightest perturbation can render a program totally worthless.
These difficulties often frustrate novice students. Instructors move to explain
more advanced subjects, while a group of students is still struggling with
basic syntax.
Again Jenkins \cite{jenkins2001motivation, jenkins2002difficulty} notes that
many students expect the course to be difficult and come with the notion that
they will have to struggle, others may have a stereotyped image of a programmer,
these beliefs can negatively affect their initial motivation and expectations.
During the course, novice programmers experience many emotions,
for instance frustration and
confusion when they cannot find a bug in a program, but also joy when they
successfully run a challenging program for the first time. They can also become
bored if they found the assignments too repetitive or too easy.

Learning to program is a difficult task, where emotions play a significant
role. Research has gone far to understand the role of emotion in learning, both
in the process and the outcome; for instance, in e-learning
\cite{kort2001affective, rossin2009effects}
and in programming  \cite{rodrigo2009affective, jenkins2001motivation ,
bosch2013emotions, khan2007mood}. In these works the emotion of flow/engagement
is regarded as the ideal affective state in which
students tend to be most capable of acquiring meaningful information through the
learning process. Flow is defined by Csikszentmihalyi \cite{csikszentmihalyi1990flow}
as a mental state in which a person performing an activity is fully immersed in a feeling of energized focus, full involvement, and enjoyment. Engagement is
also a positive affect where students are thought to be more involved
behaviorally, intellectually, and emotionally in their learning tasks
\cite{bangert2002teacher}.

>>>>

\section{Introduction}
Programming courses are often regarded as difficult
\cite{robins2003learning, lahtinen2005study,munson2018models}
and there is a common conception, backed by data, that they often have a high
failure rate \cite{bennedsen2007failure}.
Jenkins \cite{jenkins2001motivation} argues that this high rate of failure has
multiple factors:
\begin{itemize}
\item There is a deep relation with the expectations, attitudes,
and previous experiences of the teaching staff and students.
\item The nature of the subject that involves the
learning of new abstract constructs, syntax and tools.
\item Groups are often heterogeneous and thus it is difficult to design courses
that are beneficial for everyone. Many students begin to program when they
are at their first year of university which means they are tackling a totally
new topic that does not respond to their habitual study approaches.
\end{itemize}

Dijkstra \cite{dijkstra1989cruelty} argues that the subject of programming is
very problem-solving intensive, which means that a high precision is required
since even the slightest deviation from syntax can render a program
totally worthless. These are hurdles that often frustrate novice
students. This is aggravated by the fact that the time devoted to a
particular topic is fixed, and when it is exhausted, instructors move to explain
more advanced subjects, while a group of students might still be struggling with
basic syntax.
Again Jenkins \cite{jenkins2001motivation, jenkins2002difficulty} notes that
many students expect the course to be difficult and come with the notion that
they will have to struggle; others may have a stereotyped image of a programmer,
and these beliefs can negatively affect their initial motivation and expectations.
During the duration of the course and its delivery, novice programmers
have been proved to experience  emotions such as frustration and
confusion when a bug in a program eludes them, but also joy when they
successfully run a challenging program for the first time. They can also become
bored if they found the assignments too repetitive or too easy.

Learning to program is a difficult task, and similarly to other
learning tasks, emotions play a significant
role. Research has gone far to understand the role of emotion in learning, both
in the process and the outcome; for instance, in e-learning
\cite{kort2001affective, rossin2009effects}
and in programming  \cite{rodrigo2009affective, jenkins2001motivation ,
bosch2013emotions, khan2007mood}. In these works the emotion of flow/engagement
is regarded as the ideal affective state in which
students tend to be most capable of acquiring meaningful information through the
learning process. Flow is defined by Csikszentmihalyi \cite{csikszentmihalyi1990flow}
as a mental state in which a person performing an activity is fully immersed in a feeling of energized focus, full involvement, and enjoyment. Engagement is
also a positive affect where students are thought to be more involved
behaviorally, intellectually, and emotionally in their learning tasks
\cite{bangert2002teacher}.
